---
title: "02: Multiple species"
author: "Fallert, S. and Cabral, J.S."
output:
    rmarkdown::html_vignette:
        code_folding: show
vignette: >
  %\VignetteIndexEntry{02: Multiple species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

We are going to expand the example from the previous vignette to include multiple species.

# Setting up the simulation
As previously, we start by loading the packages and creating the landscape.
```{r setup, fig.cap = "Figure 1: The habitat quality of the example landscape. Note: higher value = better habitat quality"}
library(metaRange)
library(terra)

# find the file
raster_file <- system.file("ex/elev.tif", package = "terra")

# load it
r <- rast(raster_file)

# scale it
r <- scale(r, center = FALSE, scale = TRUE)

r <- rep(r, 10)
landscape <- sds(r)
names(landscape) <- c("habitat_quality")

# plot the first layer of the landscape
plot(landscape[["habitat_quality"]][[1]], main = "Habitat quality")
```

As before, we create a simulation environment and add the landscape to it.
```{r create_simulation}
sim <- create_simulation(
    source_environment = landscape,
    ID = "example_simulation",
    seed = 1)
```

# Adding more species to the simulation
Instead of only adding one species to the simulation, we can just repeat the  ``add_species()` call to add more.
```{r add_species}
species_to_add <- c("species_1", "species_2")
for (species in species_to_add) {
    sim$add_species(name = species)
}
```

If you are at any point wondering which, or how many species are in the simulation, you can use the `species_names()` method.
```{r species_names}
sim$species_names()
```

# Adding traits to (multiple) species
The `add_traits()` method is able to add (multiple) traits to multiple species at once, which is useful when setting up a large number of species with the same traits.
So instead of only specifying one `species` argument, we can specify a vector of species names, all of which will get the same traits.
```{r add_trait}
sim$add_traits(
    species = c("species_1", "species_2"),
    population_level = TRUE,
    abundance = 100
)
```

If we would want to add a trait to all species in the simulation, without having to type their names, we could use the already mentined `species_names()` method to get a vector of all species names and then use that as the `species` argument.
```{r add_trait_2}
sim$add_traits(
    species = sim$species_names(),
    population_level = TRUE,
    reproduction_rate = 0.5,
    carrying_capacity = 1000
)
```

Since we only have the two species in the simulation this would be equivalent to the previous call.


# Adding processes
Until now, we have two species in the simulation that are virtually identical.
In order to make them behave differently, we can add different processes to them.

## Reproduction
We are going to use the same reproduction process as in the previous vignette, but we will vary the influence of the habitat quality.
For species 1, we will let the carrying capacity depend on the habitat quality, while for species 2, it is going to influence the reproduction rate.
```{r Reproduction}
sim$add_process(
    species = "species_1",
    process_name = "reproduction",
    process_fun = function() {
        ricker_reproduction_model(
            self$traits$abundance,
            self$traits$reproduction_rate,
            self$traits$carrying_capacity * self$sim$environment$current$habitat_quality
        )
        print(
            paste0(self$name, " mean abundance: ", mean(self$traits$abundance))
        )
    },
    execution_priority = 1
)
```

```{r Reproduction2}
sim$add_process(
    species = "species_2",
    process_name = "reproduction",
    process_fun = function() {
    self$traits$abundance <-
        ricker_reproduction_model(
            self$traits$abundance,
            self$traits$reproduction_rate * self$sim$environment$current$habitat_quality,
            self$traits$carrying_capacity
        )
        print(
            paste0(self$name, " mean abundance: ", mean(self$traits$abundance))
        )
    },
    execution_priority = 1
)
```

# Executing the simulation
Now we can execute the simulation again and compare the results.
```{r run_simulation}
set_verbosity(0)
sim$begin()
```

```{r resA, fig.cap = "Figure 2: The resulting abundance distribution of species 1 after 10 simulation time steps."}
# define a nice color palette
plot_cols <- hcl.colors(100, "BluYl", rev = TRUE)
plot(
    sim,
    obj = "species_1",
    name = "abundance",
    main = "Species 1: abundance",
    col = plot_cols
)
```


```{r resB, fig.cap = "Figure 3: The resulting abundance distribution of species 2 after 10 simulation time steps."}
plot(
    sim,
    obj = "species_2",
    name = "abundance",
    main = "Species 2: abundance",
    col = plot_cols
)
```
