---
title: "02: Multiple species"
author: "Fallert, S. and Cabral, J.S."
output:
    rmarkdown::html_vignette:
        code_folding: show
vignette: >
  %\VignetteIndexEntry{02: Multiple species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

We are going to expand the example from the previous vignette to include multiple species.

# Setting up the simulation
As previously, we start by loading the packages and creating the landscape.
```{r setup, fig.cap = "Figure 1: The habitat quality of the example landscape. Note: higher value = better habitat quality"}
library(metaRange)
library(terra)

# find the file
raster_file <- system.file("ex/elev.tif", package = "terra")

# load it
r <- rast(raster_file)

# scale it
r <- scale(r, center = FALSE, scale = TRUE)

r <- rep(r, 10)
landscape <- sds(r)
names(landscape) <- c("habitat_quality")

# plot the first layer of the landscape
plot(landscape[["habitat_quality"]][[1]], main = "Habitat quality")
```

As before, we create a simulation and add the landscape to it.
```{r create_simulation}
sim <- create_simulation(
    source_environment = landscape,
    ID = "example_simulation",
    seed = 1
)
```

# Adding more species to the simulation
Instead of only adding one species to the simulation, we can just repeat the  ``add_species()` call to add more.
```{r add_species}
species_to_add <- c("species_1", "species_2")
for (species in species_to_add) {
    sim$add_species(name = species)
}
```

If you are at any point wondering which, or how many species are in the simulation, you can use the `species_names()` method.
```{r species_names}
sim$species_names()
```

# Adding traits to (multiple) species
The `add_traits()` method is able to add (multiple) traits to multiple species at once, which is useful when setting up a large number of species with the same traits.
So instead of only specifying one `species` argument, we can specify a vector of species names, all of which will get the same traits.
Here, we set the initial abundence of the species to be proportional to the habitat quality.
```{r add_trait}
sim$add_traits(
    species = c("species_1", "species_2"),
    population_level = TRUE,
    abundance = 200 * terra::as.matrix(
        sim$environment$sourceSDS[["habitat_quality"]][[1]],
        wide = TRUE)
)
```
```{r initAbundance, fig.cap = "Figure 2: The initial abundance of the species."}
# define a nice color palette
plot_cols <- hcl.colors(100, "BluYl", rev = TRUE)
plot(
    sim,
    obj = "species_1",
    name = "abundance",
    main = "Initial abundance",
    col = plot_cols
)
```

If we would want to add a trait to all species in the simulation, without having to type their names, we could use the already mentioned `species_names()` method to get a vector of all species names and then use that as the `species` argument.
```{r add_trait_2}
sim$add_traits(
    species = sim$species_names(),
    population_level = TRUE,
    reproduction_rate = 1.5,
    carrying_capacity = 1000,
    allee_threshold = 150
)
```

Since we only have the two species in the simulation this would be equivalent to the previous call.


# Adding processes
Until now, we have two species in the simulation that are virtually identical.
In order to make them behave differently, we can add different processes to them.

## Reproduction
We are going to use two different reproduction models and we will let the habitat quality influence the carrying capacity.
In the case of species 1, we will use the same reproduction model from the previous vignette `ricker_reproduction_model`.
```{r Reproduction}
sim$add_process(
    species = "species_1",
    process_name = "reproduction",
    process_fun = function() {
        ricker_reproduction_model(
            self$traits$abundance,
            self$traits$reproduction_rate,
            self$traits$carrying_capacity * self$sim$environment$current$habitat_quality
        )
        print(
            paste0(self$name, " mean abundance: ", mean(self$traits$abundance))
        )
    },
    execution_priority = 1
)
```
In the case of species 2, we will use a Ricker model with additional allee effects `ricker_allee_reproduction_model`
as described in: Cabral, J.S. and Schurr, F.M. (2010) [Ref. 1].
This means the populations that are smaller than the allee threshold will have a negative per-capacity reproduction rate.
The Allee effects is also known as "depensation" or "negative density dependence" and can describe
multiple different mechanisms that lead to lower reproduction rates at small population sizes as for example
difficulties finding a mate, or increased predation pressure (see: Liermann and Hilborn, 2001) [Ref. 2].

```{r Reproduction2}
sim$add_process(
    species = "species_2",
    process_name = "reproduction",
    process_fun = function() {
        self$traits$abundance <-
            ricker_allee_reproduction_model(
                self$traits$abundance,
                self$traits$reproduction_rate,
                self$traits$carrying_capacity * self$sim$environment$current$habitat_quality,
                self$traits$allee_threshold
            )
        print(
            paste0(self$name, " mean abundance: ", mean(self$traits$abundance))
        )
    },
    execution_priority = 1
)
```

# Executing the simulation
Now we can execute the simulation again and compare the results.
```{r run_simulation}
set_verbosity(0)
sim$begin()
```

```{r resA, fig.cap = "Figure 3: The resulting abundance distribution of species 1 after 10 simulation time steps."}
plot(
    sim,
    obj = "species_1",
    name = "abundance",
    main = "Species 1: abundance",
    col = plot_cols
)
```


```{r resB, fig.cap = "Figure 4: The resulting abundance distribution of species 2 after 10 simulation time steps."}
plot(
    sim,
    obj = "species_2",
    name = "abundance",
    main = "Species 2: abundance",
    col = plot_cols
)
```

Note how the populations of species 2 that are smaller than the allee threshold going extinct.


# References:
(1) Cabral, J.S. and Schurr, F.M. (2010) Estimating demographic models for the range dynamics of plant species. Global Ecology and Biogeography, 19, 85--97. doi:10.1111/j.1466-8238.2009.00492.x

(2) Liermann, and Hilborn, (2001), Depensation: evidence, models and implications. Fish and Fisheries, 2: 33--58. doi:10.1046/j.1467-2979.2001.00029.x
