---
title: "03: Niche suitability"
author: "Fallert, S. and Cabral, J.S."
output:
    rmarkdown::html_vignette:
        code_folding: show
vignette: >
  %\VignetteIndexEntry{03: Niche suitability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```


How environmental conditions influence the distribution of species is a central study question in ecology.
In this tutorial we will show how to implement a simple niche suitability model in metaRange.
For this, we will first create a landscape with two environmental variables (temperature and precipitation) and then add two species to it, that have different environmental preferences.

# Setup
First we load the packages and create the example landscape.

```{r create_landscape}
library(metaRange)
library(terra)
set_verbosity(2)

# find the example raster file
raster_file <- system.file("ex/elev.tif", package = "terra")

# load it
r <- rast(raster_file)

# adjust the values
temperature <- scale(r, center = FALSE, scale = TRUE) * 10 + 273.15
precipitation <- r * 2
```


Now we can (again) turn the raster into an `SDS` with one layer per time step.

```{r}
temperature <- rep(temperature, 10)
precipitation <- rep(precipitation, 10)
landscape <- sds(temperature, precipitation)
names(landscape) <- c("temperature", "precipitation")
```



```{r landscape_01, fig.cap = "Figure 1: The temperature of the example landscape. Only the first layer of 10 identical ones is shown."}
terra::plot(
    landscape$temperature[[1]],
    col = hcl.colors(100, "RdYlBu", rev = TRUE),
    main = "Temperature [K]"
)
```
```{r landscape_02, fig.cap = "Figure 2: The precipitation of the example landscape. Only the first layer of 10 identical ones is shown."}
terra::plot(
    landscape$precipitation[[1]],
    col = hcl.colors(100, "Earth"),
    main = "Precipitation [mm]"
)
```

# Creating the simulation and species
```{r add_species}
sim <- create_simulation(
    source_environment = landscape,
    ID = "example_simulation",
    seed = 1
)
sim$add_species(name = "species_1")
sim$add_species(name = "species_2")
```

# Adding traits to species
To start, we add the same traits as in the previous tutorials and also add a trait called `climate_suitability`, where we will store the information about how suitable the environment in each cell is for the population that lives there.
```{r add_trait}
sim$add_traits(
    species = c("species_1", "species_2"),
    population_level = TRUE,
    abundance = 500,
    climate_suitability = 1,
    reproduction_rate = 0.3,
    carrying_capacity = 1000
)
```

Contrary to the above, some traits may not require to be stored at the population level.
In this example, this could be the case for the environmental preferences of a species.
If we assume that they are a property of the species as a whole (i.e. the same for all populations), we can set the parameter `population_level` to `FALSE` and the traits will be added as they are, without further processing.

In order to make the simulation more interesting, we can give both species different environmental preferences for the two environmental variables in the simulation environment (temperature & precipitation).

Note that the names of the traits are arbitrary and can be chosen by the user and that there is no predetermined connection between e.g. "min_temperature" and the temperature variable in the environment.
It is the responsibility of the species processes to access the correct traits and use them in a sensible way, which is why meaningful trait names are important.
```{r add_trait_3}
sim$add_traits(
    species = "species_1",
    population_level = FALSE,
    max_temperature = 300, # Kelvin
    optimal_temperature = 288,
    min_temperature = 280,
    max_precipitation = 1000, # mm
    optimal_precipitation = 700,
    min_precipitation = 200
)
sim$add_traits(
    species = "species_2",
    population_level = FALSE,
    max_temperature = 290,
    optimal_temperature = 285,
    min_temperature = 270,
    max_precipitation = 1000,
    optimal_precipitation = 500,
    min_precipitation = 0
)
```


# Adding processes

## Calculate the suitability
To calculate the suitability, we use the metaRange function `calculate_suitability()` that was adapted from a formula published by [Yin et al. in 1995](https://doi.org/10.1016/0168-1923(95)02236-Q) [Ref. 1] and simplified by [Yan and Hunt in 1999 (eq:4)](https://doi.org/10.1006/anbo.1999.0955) [Ref. 2].
The formula takes the three cardinal values of an environmental niche (minimum tolerable value, optimal vale and maximum tolerable value) and constructs a suitability curve based on a beta distribution.

```{r calculate_suitability, fig.cap="Figure 3: Example suitability curve for the temperature niche of species 2."}
min_value <- 270
opt_value <- 285
max_value <- 290
x <- seq(min_value, max_value, length.out = 100)
y <- calculate_suitability(max_value, opt_value, min_value, x)
plot(x, y, type = "l", xlab = "Temperature [K]", ylab = "Suitability")
```


In the following code we add a process to both species that calculates the environmental suitability for precipitation and temperature and then multiplies the values to create a joint suitability over the two niches.
Note that one could also define a custom function to calculate the suitability, if this built-in function does not adequately describe the ecology of the target species.

## Suitability
```{r Suitability}
sim$add_process(
    species = c("species_1", "species_2"),
    process_name = "calculate_suitability",
    process_fun = function() {
        self$traits$climate_suitability <-
            calculate_suitability(
                self$traits$max_temperature,
                self$traits$optimal_temperature,
                self$traits$min_temperature,
                self$sim$environment$current$temperature
            ) *
            calculate_suitability(
                self$traits$max_precipitation,
                self$traits$optimal_precipitation,
                self$traits$min_precipitation,
                self$sim$environment$current$precipitation
            )
    },
    execution_priority = 1
)
```

## Reproduction
As in the previous tutorials, we use a Ricker reproduction model to calculate the new abundance of the species, but this time we let the carrying capacity and the reproduction rate depend on the niche suitability of the environment.
```{r Reproduction}
sim$add_process(
    species = c("species_1", "species_2"),
    process_name = "reproduction",
    process_fun = function() {
        self$traits$abundance <-
            ricker_reproduction_model(
                self$traits$abundance,
                self$traits$reproduction_rate * self$traits$climate_suitability,
                self$traits$carrying_capacity * self$traits$climate_suitability
            )
    },
    execution_priority = 2
)
```

# Results
Now, we can execute the simulation and compare the results.
```{r run_simulation}
set_verbosity(1)
sim$begin()
```

```{r resA, fig.cap = "Figure 6: The resulting abundance distribution of species 1 after 10 simulation time steps."}
# define a nice color palette
plot_cols <- hcl.colors(100, "Purple-Yellow", rev = TRUE)
plot(
    sim,
    obj = "species_1",
    name = "abundance",
    main = "Species 1: abundance",
    col = plot_cols
)
```


```{r resB, fig.cap = "Figure 7: The resulting abundance distribution of species 2 after 10 simulation time steps."}
plot(
    sim$species_2,
    trait = "abundance",
    main = "Species 2: abundance",
    col = plot_cols
)
```


# References
(1) Yin, X., Kropff, M.J., McLaren, G., Visperas, R.M., (1995) A nonlinear model for crop development as a function of temperature, Agricultural and Forest Meteorology, Volume 77, Issues 1-2, Pages 1--16, doi:10.1016/0168-1923(95)02236-Q

(2) Yan, W., Hunt, L.A. (1999) An Equation for Modelling the Temperature Response of Plants using only the Cardinal Temperatures, Annals of Botany, Volume 84, Issue 5, Pages 607--614, ISSN 0305-7364, doi:10.1006/anbo.1999.0955
