---
title: "10: Saving results"
author: "Fallert, S. and Cabral, J.S."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10: Saving results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

At the end of a simulation you will most likely want to save the results to perform further analysis in the future without having to repeat the whole simulation.
This can be done through the function `save_species()`.
It takes as input a (metaRange) species object and will by default saves all its traits, or only a subset, if the user list specific trait names.

If we take the example simulation from the first vignette:

```{r setup}
library(metaRange)
library(terra)
raster_file <- system.file("ex/elev.tif", package = "terra")
r <- rast(raster_file)
r <- scale(r, center = FALSE, scale = TRUE)
r <- rep(r, 10)
landscape <- sds(r)
names(landscape) <- "habitat_quality"

sim <- create_simulation(landscape)
sim$add_species("species_1")

sim$add_traits(
    species = "species_1",
    population_level = TRUE,
    abundance = 100,
    reproduction_rate = 0.5,
    carrying_capacity = 1000
)

sim$add_process(
    species = "species_1",
    process_name = "reproduction",
    process_fun = function() {
        ricker_reproduction_model(
            self$traits$abundance,
            self$traits$reproduction_rate,
            self$traits$carrying_capacity * self$sim$environment$current$habitat_quality
        )
    },
    execution_priority = 1
)

sim$begin()
```


We can now call `save_species()` and supply the species as input and also a path to a folder / directory where the results should be stored.
```{r save_1}
directory_name <- tempdir()

result_paths <- save_species(
    sim$species_1,
    path = directory_name
)

basename(result_paths)
```

Because we don't want to generate useless files when building this document we are now going to delete these files again:
```{r unlink1}
unlink(result_paths)
```

# Saving a time series of traits
Another important aspect of the analysis of the simulation results is the change over time.
So how can we save a time series of the traits and not just the last time step?
We can leverage the option to add global processes and just include the call to `save_species()` there.

Here we will illustrate this with an example that includes two species:

```{r save_2}
sim <- create_simulation(landscape)#
sim$set_time_layer_mapping(c(1, 1, 1))
sim$add_species(c("species_1", "species_2"))

sim$add_traits(
    species = c("species_1", "species_2"),
    population_level = TRUE,
    abundance = 100,
    reproduction_rate = 0.5,
    carrying_capacity = 1000
)

sim$add_process(
    species = c("species_1", "species_2"),
    process_name = "reproduction",
    process_fun = function() {
        ricker_reproduction_model(
            self$traits$abundance,
            self$traits$reproduction_rate,
            self$traits$carrying_capacity * self$sim$environment$current$habitat_quality
        )
    },
    execution_priority = 1
)

sim$add_globals(
    result_paths = c(),
    save_dir = tempdir()
)

sim$add_process(
    #  "species" is NULL
    process_name = "saving_traits",
    process_fun = function() {
        # loop over all species
        for (species in self$species_names()){
            result_paths <- save_species(
                # pass the species object
                self[[species]],
                # specify traits we want to save
                traits = "abundance",
                # a prefix for each time step
                prefix = paste0(self$get_current_time_step(), "-"),
                # where should it be saved
                path = self$globals$save_dir
            )
            print(c("saved: ", basename(result_paths)))
            self$globals$result_paths <- c(
                self$globals$result_paths,
                result_paths
            )
        }
    },
    execution_priority = 2
)

sim$begin()
```

Now delete the files again.
Don't do this if you actually want to keep the files.
```{r unlink2}
unlink(sim$globals$result_paths)
```
